\documentclass[../main.tex]{subfiles}


\begin{document}

\section{Conclusion}
\label{sec:conclusion}

This report walked the full alignment pipeline for SmolLM2-360M on the IT job-postings domain, covering supervised fine-tuning, RLAIF-based preference dataset construction, and DPO training at four KL-coefficient values.

Supervised fine-tuning transformed the domain-adapted text completer from Mini-Assignment~1 into a structured instruction follower reaching a validation perplexity of 4.46 with only 2.34\% of parameters trainable. DPO then refined that policy against preferences using a cross-judge protocol that keeps the RLAIF and evaluation roles assigned to model families distinct from one another and from the ETL teacher. The deliverable policy, DPO-$\beta$=0.10, defeats the base model on 18 of 18 consistent judge verdicts (90\% AB-BA agreement) and sweeps SFT on every consistent close-pair verdict (8-0, zero SFT wins). A four-point beta sensitivity probe identifies $\beta = 0.10$ as the strongest candidate among the tested configurations: perplexity and win-rate both degrade below that value, with $\beta = 0.05$ producing a 47\% perplexity cliff and a 40\% drop in mean output length. The close-pair AB-BA agreement rates documented in Section~\ref{sec:limitations} prevent a statistically robust ranking between the three in-range DPO variants; $\beta = 0.10$ is the recommended deliverable on best-available-evidence grounds rather than as a proven optimum.

Three methodological lessons emerged from the iteration history. First, inference-time decoder kwargs are load-bearing: dropping \texttt{repetition\_penalty=1.3} from the evaluation generation function, a parameter that solved the identical failure mode in Mini-Assignment~1, produced catastrophic token-loop collapses that the training-side metrics could not detect. Second, the calibration sample used to select the evaluation judge was undersized for the question it was asked: a 10-call probe cannot distinguish a 80\% from a 50\% AB-BA agreement rate, and production scale at 600 calls surfaced a structural discrimination ceiling on close pairs. Third, preference-data volume was not the bottleneck: a 6.5$\times$ scaling of preference triples did not move the over-fit signature, the judge floor, or the deployment-side win-rate; the binding constraint was preference-signal quality and judge capacity.

The aligned checkpoint, DPO-b01, is the input to the Final Project, where it will be integrated into a downstream recruiter-assistance system.

\end{document}

\end{document}
